---
title: "NYC Car Crash Analysis"
subtitle: "Final Project Presentation"
author: "Kevin Lei, Micah Kepe, Zachary Kepe, Giulia Costantini"
format:
  revealjs:
    slide-number: true
    chalkboard: 
      buttons: false
    preview-links: auto
    footer: "STAT 405 Group 3 - Micah Kepe, Kevin Lei, Zachary Kepe, Giulia Costantini"
server: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

# Load required packages
library(reshape2)
library(tidyverse)
library(lubridate)
library(ggplot2)
library(plotly)
library(chron)
library(reshape2)
library(RColorBrewer)
library(scales)
library(rpart)
library(rpart.plot)
library(sjPlot)
library(DBI)
library(RSQLite)
library(sqldf)
library(kableExtra)

# TODO: Load the data

```

## Introduction {.smaller auto-animate="true"}

- **Objective**: Analyze car crash data in NYC from 2012 to 2024
- **Data Description**: 
  - Date, time, location of crashes
  - Number of people injured or killed in each crash
- **Motivation**:
  - Identify trends and patterns
  - Provide insights for improving road safety

## Data Cleaning {.smaller auto-animate="true"}

- Removed rows with missing or invalid values
- Converted date and time columns to standard formats
- Replaced empty strings with "UNKNOWN" for street names
- Dropped redundant location column
- Removed remaining rows with NA values

## Key Findings {.smaller auto-animate="true"}

- Brooklyn has the highest percentage of crashes per capita
- Crash density is highest in central areas with heavy traffic
- Crashes peak during evening rush hours
- Driver inattention/distraction is the leading cause of crashes
- Fatal crashes are more prominent during evening and night
- Top 5 contributing factors have been decreasing since 2020

## Crash Percentage by Borough {auto-animate="true"}

```{r}
# Insert code
```

## Crash Density Heatmap {auto-animate="true"}

```{r}
# Insert code
```

## Crashes by Time of Day {auto-animate="true"}

```{r}
# Insert code
```

## Contributing Factors {auto-animate="true"}

```{r}
# Insert code
```

## Trends of Contributing Factors {auto-animate="true"}

```{r}
# Insert code
```

## Killer Plot {auto-animate="true"}

<iframe src="https://micahkepe.shinyapps.io/NYC-Crashes/" width="100%" height="650px"></iframe>

## Predictive Modeling {.smaller auto-animate="true"}

## Conclusion {.smaller auto-animate="true"}

- Identified patterns and insights into car crashes in NYC
- Evening rush hours and driver inattention are major risk factors
- Geography and time of day play significant roles
- COVID-19 pandemic has impacted crash trends
- Findings can inform targeted interventions and policies for improving road safety

## Thank You! {.smaller auto-animate="true"}

Any questions?

:::footer
Source code available at [GitHub](https://github.com/micahkepe/stat405project)
:::





