---
title: "Group 3: Crash Analysis NYC"
author: "Kevin Lei, Micah Kepe, Zachary Kepe, Giulia Costantini"
date: ''
format:
  pdf: default
  html: default
pdf-engine: pdflatex
editor: visual
fontsize: 10pt
geometry: margin=1in
toc: true                   # add table of contents at the beginning
toc-depth: 2                # Only titles that start with # or ##
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
data <- read.csv("./data/Motor_Vehicle_Collisions_-_Crashes_20240213.csv")
as.data.frame(data)
```

#### Introduction

Our dataset contains details of motor vehicle collision occurrences reported by the NYPD from 2013 to 2023.

Collisions are only reported if a person was injured or killed, or if there was more than \$1000 in damage.

Our project seeks to determine the most important factors that contribute to potentially fatal accidents. We hope to provide value input to drivers in New York City in order to prevent fatalities.

#### Plot 1: Histogram

```{r}
borough <- table(data$BOROUGH)

table(data$BOROUGH)

names(borough) <- c("NA", "BRX", "BRN", "MAN", "QNS", "SI")

barplot(borough, las = 1,
        main = "Frequency of Car Crashes by NYC Borough (2013-2023)",
        xlab = "Borough",
        ylab = "Frequency (in hundreds of thousands)",
        yaxt = "n")

axp <- seq(0, 700000, by = 100000)
axis(2, at = axp, labels = c(0:7), las = 2)
```

```{r}
install.packages("chron")
library(chron)
times <- data$CRASH.TIME
times <- paste0(times, ":00")
times <- chron(times = times)

plot(times, data$NUMBER.OF.PERSONS.KILLED, xaxt = "n")

axp <- seq(0, 24, by = 1)
axis(1, at = axp, labels = c(0:24), las = 2)
```
